Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-01 workflow/workflow.Rmd
Updates to text.

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied based clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings by resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005]. A consensus clustering is obtained based on the levels of co-clustering of samples across the candidate clusterings. The consensus clustering is further condensed by merging similar clusters which is done by creating a hierarchy of clusters, working up the tree, and testing for differential expression between sister nodes, with nodes of insufficient DE collapsed. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.

@@

The defaults in `RSEC` are designed for input data that is the actual (normalized) count data. We are going to apply it instead to the `W` matrix of the previous step, for which we make a separate `SummarizedExperiment` object from `W`. For this reason we will choose to not use certain options in `RSEC`. In particular, we will not use the default dimensionality reduction step provided by `RSEC` since our input `W` is already in a reduced dimensional space. Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`dimReduce`) and to the construction of the hiearchy between the clusters (`dendroReduce`). We will also skip the option to merge our clusters based on the amount of differential gene expression between clusters since our input data is `W` and not the actual individual gene expression values.

@@

The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`. The first takes a low-dimensional representation of the cells and a vector of cluster labels. It fits an MST to the clusters and identifies lineages as paths through this tree. The output of `getLineages` is an object of class `SlingshotDataSet` containing all the information used to fit the tree and identify lineages. The function `getCurves` then takes this object as input and fits simultaneous principal curves to the identified lineages. These functions can be run separately, as below, or jointly by the wrapper function, `slingshot`.

@@

[SD: Mention reason for having these two figures. What do colors correspond to in second figure? Match notation for dimensions (along diagonal) in the two figures.]

@@

In the workflow, we recover a reasonable pseudotime ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `getLineages` has the option for the user to provide known end cluster(s). Here is the code to use `slingsho` in a supervised setting.

@@

After assigning the cells to lineages and ordering them within lineages, we are interested in finding genes that have non-constant expression patterns over pseudotime. We can see global trends in expression by subsetting our original dataset to the most variable genes and displaying their expression values in a heatmap, with cells ordered by pseudotime.

More formally, we are interested in testing for variable expression over pseudotime. We do this by fitting a loess curve to the data which is a flexible, nonlinear way to model the relationship between a gene's normalized expression measures and pseudotime. We then can test the null hypothesis of no change over time using the `gam` package. We implemented this model for each gene and show the top 100 genes by p-value in the heatmap below.

>>>>

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied base clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings, with the option of resampling cells and using a sequential tight clustering procedure as in Tseng and Wong (2005)[@Tseng2005]. A consensus clustering is obtained based on the levels of co-clustering of samples across the candidate clusterings. The consensus clustering is further condensed by merging similar clusters which is done by creating a hierarchy of clusters, working up the tree, and testing for differential expression between sister nodes, with nodes of insufficient DE collapsed. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.

@@

Note that the defaults in `RSEC` are designed for input data that are the actual (normalized) counts. Here, we are applying `RSEC` instead to the low-dimensional `W` matrix of the previous step, for which we make a separate `SummarizedExperiment` object. For this reason, we choose to not use certain options in `RSEC`. In particular, we do not use the default dimensionality reduction step, since our input `W` is already in a space of reduced dimension. Specifically, `RSEC` offers a dimensionality reduction option for the input to both the clustering routines (`dimReduce`) and the construction of the hiearchy between the clusters (`dendroReduce`). We also skip the option to merge our clusters based on the amount of differential gene expression between clusters since our input data is `W` and not the actual individual gene expression values.

@@

The two steps of the Slingshot algorithm are implemented in the functions `getLineages` and `getCurves`. The first takes a low-dimensional representation of the cells and a vector of cluster labels. It fits an MST to the clusters and identifies lineages as paths through this tree. The output of `getLineages` is an object of class `SlingshotDataSet` containing all the information used to fit the tree and identify lineages. The function `getCurves` then takes this object as input and fits simultaneous principal curves to the identified lineages. These functions can be run separately, as below, or jointly by the wrapper function `slingshot`.

@@

[SD: Mention reason for having these two figures, otherwise drop second figure. If keep second figure: Move it first, match notation for dimensions (along diagonal) in the two figures, and provide color legend for second figure.]

@@

In the workflow, we recover a reasonable pseudotime ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `getLineages` has the option for the user to provide known end cluster(s). Here is the code to use `slingshot` in a supervised setting.

@@

After assigning the cells to lineages and ordering them within lineages, we are interested in finding genes that have non-constant expression patterns over pseudotime. We can see global trends in expression by subsetting our original dataset to the most variable genes [SD: How many genes?] and displaying their expression values in a heatmap, with cells ordered by pseudotime.

More formally, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime. We then can test the null hypothesis of no change over time using the `gam` package. We implemented this approach for each gene and display the expression measures of the top 100 genes by p-value in the heatmap of Figure [SD: Add figure ref.].
